Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-01 introduction.tex

Many phenomena in astrophysics and cosmology require a high spatial and temporal dynamical range and there are a number of numerical techniques which have been used to model such problems.  At the current time, the most common is a gridless, fully Lagrangian method call Smoothed Particle Hydrodynamcs \citep[SPH;][]{Lucy77, SPH} which has been very successful, particularly for those situations dominated by gravity.  Its development, however, is still at a relatively early stage when compared to the effort put into Eulerian, grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.

Therefore, we would like to make use of this expertise that has been invested in Eulerian solvers; unfortunately, as originally developed, none of these schemes provide an easy way to adaptively and dynamically increase the resolution in small volumes, as required by, for example, gravitational instability.  We have instead, adopted an idea first suggested by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, which has become known as Structured Adaptive Mesh Refinement (SAMR).  The principle is to adaptively add (and modify) additional, finer meshes, or patches, over regions which require higher resolution.  Since each patch has a simple geometry (here, Cartesian) the hyperbolic fluid-dynamical equations can be solved on that patch with a well-developed `off-the-shelf' numerical method.  In addition, it is relatively straightforward to add other advanced physics (e.g. MHD, radiation transport, etc.). This paper is devoted to describing how we have implemented this idea for astrophysics, including the addition of comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, MHD, radiation transport, star formation and a range of other physical effects.

There have been many other numerical methods described in the astronomical literature which contain elements of this idea, or have a similar aim.  For example, an N-body solver \cite{Villumsen89} used non-adaptive meshing to increase the resolution in pre-selected regions.  This static approach, applied to hydrodynamics, has been used extensively \citep[e.g.,]{Ruffert94, Anninos94}.  Adding adaptivity is a more recent enhancement, and there are now a number of N-body codes which possess this feature, both with and without  hydrodynamics  \citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97, Truelove98, flash_method, MLAPM01,  Yahagi01, RAMSES, Quilis04, Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}.  Of these, perhaps the most comparable and widely used are FLASH \citep{flash_method}, which uses grid blocks of fixed size, and RAMSES \citep{RAMSES} and ART \citep{ART97}, both of which refine individual cells.   It is also possible to deform the grid to obtain high resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a few grid codes have adopted an unstructured approach based on a moving Voronoi mesh \citep{Arepo10, Tess11}.

\enzo\ is a structured adaptive mesh refinement (SAMR) code that was originally developed for cosmological hydrodynamics and has been used on a wide variety of problems.  It has been expanded as a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily expanded.  Although many of the components of the code have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04}, there has been no systematic and complete description of the code.  In this paper we provide that description, filling in many gaps and showing the result of the code on a wide variety of test problems.

The \enzo\ code has been widely used over the last two decades in a wide variety of problems, and more than 100 peer-reviewed papers have been published based on results obtained with the code.  The code has previously been used to study a wide range of astrophysical systems, including X-ray clusters \citep{Loken02}, the ISM \citep{Slyz05}, the intergalactic medium \citep{Fang01}, cooling flows \citep{Li12}, turbulence \citep{Kritsuk04}, and the formation of the first stars \citep{ABN02}.

@@

The structure of the code is as follows.  In section~\ref{sec.overview}, we first provide a top-level overview of the code method and structure.  This is designed to give a first broad-brush picture of the equations we aim to solve and the methods used to solve them.  Next, in section~\ref{sec.methods}, we describe the methods we use in detail, reserving some of the longer descriptions of particular components for the appendix in order not to interrupt the flow of the paper.  The code tests are described in section~\ref{sec.tests} and the parallelism strategy and scaling results in section~\ref{sec.performance}.

>>>>

Many phenomena in astrophysics and cosmology require a high spatial and temporal dynamical range, and there are a number of numerical techniques that have been used to model such problems.  At the current time, the most common is a gridless, fully Lagrangian method call Smoothed Particle Hydrodynamcs \citep[SPH;][]{Lucy77, SPH} which has been very successful -- particularly for those situations dominated by gravity, where systems evolve over large spatial and temporal dynamic ranges.  Its development, however, is still at a relatively early stage when compared to the effort put into Eulerian, grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.

Therefore, we would like to make use of the expertise that has been invested in Eulerian solvers; unfortunately, as originally developed, none of these schemes provide an easy way to adaptively and dynamically increase the spatial and temporal resolution in small volumes, as required by, for example, gravitational instability.  We have instead adopted an idea first suggested by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, which has become known as Structured Adaptive Mesh Refinement (SAMR).  The principle is to adaptively add and modify additional, finer meshes (``patches'') over regions that require higher resolution.  Since each patch has a simple geometry (Cartesian, in our case) the hyperbolic fluid-dynamical equations can be solved on that patch with a well-developed `off-the-shelf' numerical method.  In addition, it is relatively straightforward to add other advanced physics (e.g. magnetohydrodynamics, radiation transport, chemistry, etc.). This paper is devoted to describing how we have implemented this idea for astrophysics, including the addition of comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, MHD, radiation transport, star formation and a range of other physical effects.

There have been many other numerical methods described in the astronomical literature which contain elements of this idea, or have a similar aim.  For example, an N-body solver \citep{Villumsen89} used non-adaptive meshing to increase the resolution in pre-selected regions.  This static approach, applied to hydrodynamics, has been used extensively \citep[e.g.,][]{Ruffert94, Anninos94}.  Adding adaptivity is a more recent enhancement, and there are now a number of N-body codes that possess this feature, both with and without  hydrodynamics  \citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97, Truelove98, flash_method, MLAPM01,  Yahagi01, RAMSES, Quilis04, Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}.  Of these, perhaps the most comparable and widely used are FLASH \citep{flash_method}, which uses grid blocks of fixed size, and RAMSES \citep{RAMSES} and ART \citep{ART97}, both of which refine individual cells.   It is also possible to deform the grid to obtain high resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a few grid codes have adopted an unstructured approach based on a moving Voronoi mesh \citep{Arepo10, Tess11}.

\enzo\ is a structured adaptive mesh refinement (SAMR) code that was originally developed for cosmological hydrodynamics and has been used on a wide variety of problems.  It has been expanded as a general tool for astrophysical fluid dynamics and is intended to be efficient, accurate and easily expanded.  Although many of the components of the code have been described in previous publications \citep{1995CoPhC..89..149B, BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01, Oshea04, 2007arXiv0705.1556N}, there has been no systematic and complete description of the code.  In this paper we provide that description, filling in many gaps and showing the code's performance for a wide variety of test problems.

The \enzo\ code has been extensively used over the last two decades in a wide variety of problems, and more than 100 peer-reviewed papers have been published based on results obtained with the code.  The code has previously been used to study a wide range of astrophysical systems, including galaxy clusters \citep{Loken02}, the interstellar medium \citep{Slyz05}, the intergalactic medium \citep{Fang01}, cooling flows \citep{Li12}, turbulence \citep{Kritsuk04}, and the formation of the first stars \citep{ABN02}.

@@

The structure of the code is as follows.  In Section~\ref{sec.overview}, we first provide a top-level overview of the code method and structure.  This is designed to give a first broad-brush picture of the equations we aim to solve and the methods used to solve them.  Next, in Section~\ref{sec.methods}, we describe the methods we use in detail, reserving some of the longer descriptions of particular components for the appendix in order not to interrupt the flow of the paper.  The \enzo\ testing framework and code tests are described in Section~\ref{sec.tests}.  The parallelism strategy and scaling results are described in Section~\ref{sec.parallel}.  Finally, we discuss the code's development methodology (which is as far as we know unique in the astrophysics community) in Section~\ref{sec.development}.

### 2013-05-01 numerical-analysis.tex

Analyzing the results of simulations in detail requires both tools to ask
sophisticated questions of data, and the ability to process vast quantities of
data at high-cadence.  As simulations grow in size and complexity, storing data
for post-processing becomes simply intractable.  To cope with this, we have
instrumented Enzo with the ability to conduct analysis during the course of a
simulation.  This enables analyzing with extremely high cadence (even as often
as every subcycle of the finest refinement level), as well as without
attempting to write an entire checkpoint output to disk.  The current mechanism
for conducting analysis in Enzo during the course of the simulation utilizes
the same compute resources as used by the simulation itself; this is often
referred to as \textit{in situ} visualization.  Utilizing a
dynamically-scheduled second set of computing resources, often referred to as
\textit{co-scheduled} visualization, provides greater flexibility and overall
throughput at the expense of simplicity.

We expose Enzo's mesh geometry and fluid quantities to the analysis platform
\texttt{yt} \citep{2011ApJS..192....9T, 2011arXiv1112.4482T}.  At compile time,
Enzo is (dynamically or statically) linked against the Python and NumPy
libraries necessary to create proxy objects exposing the mesh geometry,
fluid quantities and particle arrays as NumPy arrays.  This information is then
passed to a special handler inside \texttt{yt}.  \texttt{yt} interprets the
mesh and fluid information and, without saving data to disk except as
explicitly requested, constructs a native representation of the in-memory state
of the simulation that appears identical to an on-disk simulation output.
Enzo then executes a user-provided analysis script, which is able to access the
in-memory simulation object.  Once the analysis script has returned control to
Enzo, the simulation proceeds.  This process can occur at either the top of the
main ``EvolveHierarchy'' loop or at the end of a timestep at the finest level,
and the frequency with which it is called is adjustable by a run-time
parameter.  During the course of conducting analysis, this simulation is halted
until the conclusion of the analysis.

@@

At user-specified intervals, the tracer particles output their data
in a streaming format into a binary file.  The data written out
typically includes the tracer particle's unique ID, position, velocity, density,
and temperature, but is easily extensible to output any grid-based
quantity that a user requires.  This capability has been used quite
effectively in several papers, including \citep{2010ApJ...715.1575S, 2012ApJ...748...12S}.

@@

Identification of shocks and their pre- and post-shock conditions can be
accomplished through a combination of either temperature or velocity jumps with
dimensionally split or unsplit search methods.  The primary method used in
\enzo~is the dimensionally unsplit temperature jump method described in detail
in \citet{2008ApJ...689.1063S}.  We briefly outline the method below.

For every cell, we first determine whether is satisfies the following conditions
necessary to be flagged as a shock:
\begin{eqnarray}
\nabla \cdot \vec{v} < 0 \\
\nabla T \cdot \nabla S > 0\\
T_2 > T_1 \\
\rho_2 > \rho_1,
\end{eqnarray}
where $\vec{v}$ is the velocity field, $T$ is the temperature, $\rho$ is the
density, and $S=T/\rho^{\gamma-1}$ is the entropy.  An optional temperature floor
may be chosen, useful for example in cosmological simulations without a radiation
background.

>>>>

Analyzing the results of simulations in detail requires both tools to ask
sophisticated questions of data and the ability to process vast quantities of
data at high time-cadence.  As simulations grow in size and complexity, storing data
for post-processing simply becomes intractable.  To cope with this, we have
instrumented Enzo with the ability to conduct analysis during the course of a
simulation.  This enables analysis with extremely high time cadence (even as often
as every subcycle of the finest refinement level), as well as without
attempting to write an entire checkpoint output to disk.  The current mechanism
for conducting analysis in Enzo during the course of the simulation utilizes
the same compute resources as used by the simulation itself; this is often
referred to as \textit{in situ} analysis or visualization.  Utilizing a
dynamically-scheduled second set of computing resources, often referred to as
\textit{co-scheduled} analysis or visualization, provides greater flexibility and overall
throughput at the expense of simplicity.

We expose Enzo's mesh geometry and fluid quantities to the analysis platform
\texttt{yt} \citep{2011ApJS..192....9T, 2011arXiv1112.4482T}.  At compile time,
Enzo is (dynamically or statically) linked against the Python and NumPy
libraries necessary to create proxy objects exposing the mesh geometry,
fluid quantities and particle arrays as NumPy arrays.  This information is then
passed to a special handler inside \texttt{yt}.  \texttt{yt} interprets the
mesh and fluid information and, without saving data to disk except as
explicitly requested, constructs a native representation of the in-memory state
of the simulation that appears identical to an on-disk simulation output.
Enzo then executes a user-provided analysis script, which is able to access the
in-memory simulation object.  Once the analysis script has returned control to
Enzo, the simulation proceeds.  This process can occur at either the top of the
main ``EvolveHierarchy'' loop or at the end of a timestep at the finest level,
and the frequency with which it is called is adjustable by a run-time
parameter.  During the course of conducting analysis, the simulation is halted
until the conclusion of the analysis.

@@

At user-specified intervals, the tracer particles output their data
in a streaming format into a binary file.  The data written out
typically includes the tracer particle's unique ID, position, velocity, density,
and temperature, but is easily extensible to output any grid-based
quantity that a user requires.  This capability has been used quite
effectively in several papers, including \citet{2010ApJ...715.1575S}
and \citet{2012ApJ...748...12S}.

@@

Identification of shocks and their pre- and post-shock conditions can be
accomplished through a combination of either temperature or velocity jumps with
dimensionally split or unsplit search methods.  The primary method used in
\enzo~is the dimensionally unsplit temperature jump method described in detail
in \citet{2008ApJ...689.1063S}.  We briefly outline the method here.

For every cell, we first determine whether is satisfies the following conditions
necessary to be flagged as a shock:

\begin{eqnarray}
\nabla \cdot \vec{v} < 0 \\
\nabla T \cdot \nabla S > 0\\
T_2 > T_1 \\
\rho_2 > \rho_1,
\end{eqnarray}

where $\vec{v}$ is the velocity field, $T$ is the temperature, $\rho$ is the
density, and $S=T/\rho^{\gamma-1}$ is the entropy.  An optional temperature floor
may be chosen, which is useful for, e.g., cosmological simulations without a radiation
background.

### 2013-05-01 numerical-chemistry.tex

While it is often safe to assume that species (both chemical and
ionization) within a gas can be treated as being in equilibrium, in
some regimes that are found in astrophysics this is not true.  For
example, the cooling and collapse of primordial gas in Population III
star formation is dominated by molecular hydrogen, which in the
absence of dust forms via an inefficient pair of collisional processes
that depend heavily on the local population of free electrons.  As a
result, when modeling primordial star formation, it is critical to
follow the non-equilibrium evolution of the chemical species of
hydrogen, including molecular hydrogen and deuterium.

The primordial non-equilibrium chemistry routines used in \enzo\ were
first described by Abel et al. and Anninos et
al. \citep{abel97,anninos97}, but have been extended since then with
updated reaction rates and the inclusion of deuterium species
\citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12
total species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp,
and HD.  (A separate, but related, module follows the radiative
heating and cooling of the gas from atomic and molecular line
excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as
well as several models for a metagalactic UV background that heat the
gas via photoionization and discussion -- see
Section~\ref{sec.num.cooling} for more details).  Depending on user
parameters, one can follow simply atomic species (H, \Hp, He, \Hep,
\Hepp, and e$^-$), include those relevant for molecular hydrogen
formation (\HH, \HHp, and \Hm), and further include deuterium and its
species (D, \Dp, and HD).  \red{A total of 9 kinetic equations are
  solved, including 29 kinetic and radiative processes, for the 9
  species mentioned above.  See Table~\ref{table.collisional} for the
  collisional processes and Table~\ref{table.radiative} for the
  radiative processes solved.}

The chemical reaction equation network is technically challenging to
solve due to the huge range of reaction timescales involved -- the
characteristic creation and destruction timescales of the various
species and reactions can differ by many orders of magnitude.  As a
result, the set of rate equations is extremely stiff, and an explicit
scheme for integration of the rate equations can be exceptionally
costly if small enough timesteps are taken to keep the network stable.
This makes an implicit scheme much more preferable for such a set of
equations.  However, an implicit scheme typically require an iterative
procedure to converge, and for large networks (such as this one) an
implicit method can be very time-consuming, making it undesirable for
a large, three-dimensional simulation.

\enzo\ solves the rate equations using a method based on a backwards
differencing formula (BDF) in order to provide a stable and accurate
solution. This technique is optimized by taking the chemical
intermediaries \Hm and \HHp, which have large rate coefficients
and low concentrations, and grouping them into a separate category of
equations.  Due to their fast reactions these species are very
sensitive to small changes in the more abundant species.  However, due
to their low overall concentrations, they do not significantly
influence the abundance of species with higher concentrations.
Therefore, reactions involving these two species can be decoupled from
the rest of the network and treated independently.  In fact, \Hm~and
\HHp~are treated as being in equilibrium at all times, independent
of the other species and the hydrodynamic variables.  This allows a
large speedup in solution as the BDF scheme is then applied only to
the slower 7- or 10-species network (depending on whether deuterium is
included or not) on timescales closer to those required by the
hydrodynamics of the simulation.  Even so, the accuracy and stability
of the scheme is maintained by subcycling the rate solver over a
single hydrodynamic timestep.  These subcycle timesteps are determined
so that the maximum fractional change in the electron concentration is
limited to no more than $10\%$ per timestep.  Given this subcycling,
the method is not truly implicit over an entire hydrodynamical
timestep, and thus not truly a BDF method -- however, it uses the
formalism and is implicit over subcycle timesteps.

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$.  Beyond this point,
modifications the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.  Even
with these modifications, a completely correct description of the
cooling of this gas will require some form of radiation transport,
which will greatly increase the cost of the simulations.

>>>>

While it is often safe to assume that species (both chemical and
ionization) within a gas can be treated as being in equilibrium, in
some regimes that are found in astrophysics this assumption leads to
considerable error.  For example, the cooling and collapse of
primordial gas in Population III star formation is dominated by
molecular hydrogen, which in the absence of dust forms via an
inefficient pair of collisional processes that depend heavily on the
local population of free electrons.  As a result, when modeling
primordial star formation, it is critical to follow the
non-equilibrium evolution of the chemical species of hydrogen,
including molecular hydrogen and deuterium.

The primordial non-equilibrium chemistry routines used in \enzo\ were
first described by Abel et al. and Anninos et
al. \citep{abel97,anninos97}, but have been extended since then with
updated reaction rates and the inclusion of deuterium species
\citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12
total species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp,
and HD.  (A separate, but related, module follows the radiative
heating and cooling of the gas from atomic and molecular line
excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as
well as several models for a metagalactic UV background that heat the
gas via photoionization and photodissociation; see
Section~\ref{sec.num.cooling} for more details).  Depending on user
parameters, one can follow only atomic species (H, \Hp, He, \Hep,
\Hepp, and e$^-$), include those species relevant for molecular
hydrogen formation (\HH, \HHp, and \Hm), and can further include
deuterium and its species (D, \Dp, and HD).  A total of 9 kinetic
equations are solved, including 29 kinetic and radiative processes,
for the 12 species mentioned above.  See Table~\ref{table.collisional}
for the collisional processes and Table~\ref{table.radiative} for the
radiative processes solved.

The chemical reaction equation network is technically challenging to
solve due to the huge range of reaction timescales involved -- the
characteristic creation and destruction timescales of the various
species and reactions can differ by many orders of magnitude.  As a
result, the set of rate equations is extremely stiff, and an explicit
scheme for integration of the rate equations can be exceptionally
costly if small enough timesteps are taken to keep the network stable.
This makes an implicit scheme strongly preferable for such a set of
equations.  However, an implicit scheme typically require an iterative
procedure to converge, and for large networks (such as this one) an
implicit method can be very time-consuming, making it undesirable for
a large, three-dimensional simulation.

\enzo\ solves the rate equations using a method based on a backwards
differencing formula (BDF) in order to provide a stable and accurate
solution. This technique is optimized by taking the chemical
intermediaries \Hm and \HHp, which have large rate coefficients and
low concentrations, and grouping them into a separate category of
equations.  Due to their fast reactions, these species are very
sensitive to small changes in the more abundant species.  However, due
to their low overall concentrations, they do not significantly
influence the abundance of species with higher concentrations.
Therefore, reactions involving these two species can be decoupled from
the rest of the network and treated independently.  In fact, \Hm~and
\HHp~are treated as being in equilibrium at all times, independent of
the other species and the hydrodynamic variables.  This allows a large
speedup in solution as the BDF scheme is then applied only to the
slower 7- or 10-species network (depending on whether deuterium is
included or not) on timescales closer to those required by the
hydrodynamics of the simulation.  Even so, the accuracy and stability
of the scheme is maintained by subcycling the rate solver over a
single hydrodynamic timestep.  These subcycle timesteps are determined
so that the estimated fractional change in the electron concentration
is limited to no more than $10\%$ per timestep.  Given this
subcycling, the method is not truly implicit over an entire
hydrodynamical timestep, and thus not truly a BDF method -- however,
it uses the formalism and is implicit over subcycle timesteps.

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling eventually breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$ in gas of primordial composition.  Beyond this point,
modifications the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.  Even
with these modifications, a completely correct description of the
cooling of primordial gas requires some form of radiation transport,
which will greatly increase the cost of the simulations.

### 2013-05-01 numerical-conduction.tex

Enzo implements the equations of isotropic heat conduction in a manner
similar to that of \citet{2007ApJ...664..135P}, where we model the
flux of isotropic heat transport:

@@

where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\nabla T|$ is the characteristic length scale of the local
temperature gradient.  Furthermore, it is important to note that the
overall level of thermal conductivity can be strongly affected
by the presence of magnetic fields, which may strongly suppress heat
flow.  The strength and confuction of magnetic fields in the
intracluster medium is poorly understood, and it is unclear to what
extent these magnetic fields will suppress conduction.  For the
purposes of the work described in this paper, we assume that the
conductivity of the plasma can be described in terms of an effective
conductivity, which can be expressed as a fraction f$_{Sp}$ of the
Spitzer conductivity (where f$_{Sp} \leq 1.0$ are considered physical
values).

@@

Both isotropic and anisotropic thermal conduction in Enzo are treated
in an operator-split manner.  Furthermore, within the heat transport
module, transport along the x, y, and z directions are computed in a
directionally-split fashion, with heat flux along each direction
calculated at the + and - faces of the cell using the arithmetic mean
of the cell-centered temperature in cells $n$ and $n+1$ or $n-1$ and
$n$, respectively (empirically, this is more stable than taking the
geometric mean of the cell-centered temperatures).  The addition of
transport along magnetic field lines requires the calculation of
cross-terms in the temperature derivatives at cell faces, which can
result in spurious oscillations in the temperature field in regions
where the temperature gradient is strong in more than one spatial
direction.  Controlling these oscillations requires the addition of a
flux limiter for calculations of the temperature field.  In this case,
we choose the monotonized central difference flux limiter
\citep{1977JCoPh..23..263V}, which serves to maintain numerical
stability without sacrificing substantial speed or accuracy.

>>>>

\enzo\ implements the equations of isotropic heat conduction in a manner
similar to that of \citet{2007ApJ...664..135P}, where we model the
flux of isotropic heat transport:

@@

where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\nabla T|$ is the characteristic length scale of the local
temperature gradient.  Furthermore, it is important to note that the
overall level of thermal conductivity can be strongly affected
by the presence of magnetic fields, which may strongly suppress heat
flow.  The strength and distribution of magnetic fields in the
intracluster medium is poorly understood, and it is unclear to what
extent these magnetic fields will suppress conduction.  For the
purposes of the work described in this paper, we assume that the
conductivity of the plasma can be described in terms of an effective
conductivity, which can be expressed as a fraction f$_{Sp}$ of the
Spitzer conductivity (where f$_{Sp} \leq 1.0$ are considered physical
values).

@@

Both isotropic and anisotropic thermal conduction in \enzo\ are treated
in an operator-split manner.  Furthermore, within the heat transport
module, transport along the x, y, and z directions are computed in a
directionally-split fashion, with heat flux along each direction
calculated at the + and - faces of the cell using the arithmetic mean
of the cell-centered temperature in cells $n$ and $n+1$ or $n-1$ and
$n$, respectively (empirically, this is more stable than taking the
geometric mean of the cell-centered temperatures).  The addition of
transport along magnetic field lines requires the calculation of
cross-terms in the temperature derivatives at cell faces, which can
result in spurious oscillations in the temperature field in regions
where the temperature gradient is strong in more than one spatial
direction.  Controlling these oscillations requires the addition of a
flux limiter for calculations of the temperature field.  In this case,
we choose the monotonized central difference flux limiter
\citep{1977JCoPh..23..263V}, which serves to maintain numerical
stability without sacrificing substantial speed or accuracy.

### 2013-05-01 numerical-cooling.tex

As discussed in \S\ref{sec.ov.chem}, the set of reactions that
characterize a metal-free gas is simple enough to be computed in
non-equilibrium during the simulation.  Similarly, the radiative
cooling of metal-free gas is solved by directly computing the cooling
and heating rates from the following individual processes for atomic H
and He: collisional excitation and ionization, recombination,
free-free emission, Compton scattering off of the cosmic microwave
background (CMB), and photo-heating from a UV metagalactic background.
If the H$_{2}$ chemistry network is enabled, the following H$_{2}$
cooling processes are also considered: ro-vibrational transitions
\citep{1998A&A...335..403G}, heating and cooling from molecular
formation and destruction \citep{2009Sci...325..601T}, and
collision-induced emission \citep{2004MNRAS.348.1019R}.  If Deuterium
chemistry is enabled, then rotational transitions of HD
\citep{1998A&A...335..403G} are treated as well.  The radiative
cooling calculation is coupled to the update of the chemistry network
such that they both occur within the same subcycling loop.  This is
necessary in regimes where rapid cooling or change in the ionization
state occur.  In addition to the subcycle timestepping contraints
mentioned in \S\ref{sec.ov.chem}, the subcycle timestep is also not
permitted to exceed 10\% of the cooling time, $e/\dot{e}$.  A
metagalactic background affects the gas through both photo-heating and
photo-ionization.  These are treated by including redshift-dependent
photo-ionization and photo-heating rate terms in the chemistry and
cooling equations for \ion{H}{1}, \ion{He}{1}, and \ion{He}{2}.  The
black curves in Figure \ref{fig.cooling_rate} show cooling rates for a
metal-free gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ both
with and without a radiation background.  More detail on the specific
UV backgrounds present in \enzo is presented below in
\S\ref{sec.num.rad-homogeneous}.

@@

A more sophisticated method makes use of multidimensional cooling and
heating rate tables computed with the photo-ionization code,
\texttt{Cloudy} \citep{1998PASP..110..761F}.  This method, detailed in
\citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}, works by
using \texttt{Cloudy} to compute the cooling and heating rates from
the metal species only.  The primary assumption made is that of
ionization equilibrium.  The tables can have up to five dimensions,
where those dimensions are density, metallicity, electron fraction,
redshift (for an evolving metagalactic UV backgroun), and
temperature.  Tables can be created for any abundance pattern for
elements up to atomic number 30 (Zn) and for any incident radiation
field.  The cooling from H and He is calculated by the non-equilibrium
primordial chemistry and cooling solver.  The contribution from metals
to the cooling is computed within the same subcycling loop as the
coupled primordial chemistry and cooling solver.  The blue curves in
Figure \ref{fig.cooling_rate} show cooling rates calculated with this
method for a gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ and
metallicity of 0.5 Z$_{\odot}$.

\begin{figure}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/cooling_rate.eps}
  \end{center}
  \caption{Radiative cooling rates from the various cooling methods
    available in \enzo.  The black curves show cooling rates from a gas
    with primordial composition using the non-equilibrium chemistry
    network.  The blue curves are for a gas with metallicity of 0.5
    $Z_{\odot}$ computed with the Cloudy cooling method.  The solid
    black and blue lines assume collisional processes only while the
    dashed lines include photo-ionization and photo-heating from a UV
    metagalactic background at $z = 0$ with a gas number density of
    10$^{-4}$ cm$^{-3}$.  The low temperature rates shown by the
    dashed lines indicate a net heating.  The red curve is the
    tabulated cooling function of \citet{SW87} which assumes a fully
    ionized gas with metallicity of 0.5 $Z_{\odot}$.}
  \label{fig.cooling_rate}
\end{figure}

>>>>

As discussed in Section~\ref{sec.num.chemistry}, the set of reactions that
characterize a metal-free gas is simple enough to be computed in
non-equilibrium during the simulation.  Similarly, the radiative
cooling of metal-free gas is solved by directly computing the cooling
and heating rates from the following individual processes for atomic H
and He: collisional excitation and ionization, recombination,
free-free emission, Compton scattering off of the cosmic microwave
background (CMB), and photo-heating from a UV metagalactic background.
If the H$_{2}$ chemistry network is enabled, the following H$_{2}$
cooling processes are also considered: ro-vibrational transitions
\citep{1998A&A...335..403G}, heating and cooling from molecular
formation and destruction \citep{2009Sci...325..601T}, and
collision-induced emission \citep{2004MNRAS.348.1019R}.  If Deuterium
chemistry is enabled, then rotational transitions of HD
\citep{1998A&A...335..403G} are treated as well.  The radiative
cooling calculation is coupled to the update of the chemistry network
such that they both occur within the same subcycling loop.  This is
necessary in regimes where rapid cooling or change in the ionization
state occur.  In addition to the subcycle timestepping contraints
mentioned in Section~\ref{sec.num.chemistry}, the subcycle timestep is also not
permitted to exceed 10\% of the cooling time, $t_{cool} = e/\dot{e}$.  A
metagalactic background affects the gas through both photo-heating and
photo-ionization.  These are treated by including redshift-dependent
photo-ionization and photo-heating rate terms in the chemistry and
cooling equations for \ion{H}{1}, \ion{He}{1}, and \ion{He}{2}.  The
black curves in Figure \ref{fig.cooling_rate} show cooling rates for a
metal-free gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ both
with and without a radiation background.  More detail on the specific
UV backgrounds present in \enzo is presented below in
Section~\ref{sec.num.rad-homogeneous}.

@@

A more sophisticated method makes use of multidimensional cooling and
heating rate tables computed with the photo-ionization code
\texttt{Cloudy} \citep{1998PASP..110..761F}.  This method, detailed in
\citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}, works by
using \texttt{Cloudy} to compute the cooling and heating rates from
the metal species only.  The primary assumption made is that of
ionization equilibrium.  The tables can have up to five dimensions,
where those dimensions are density, metallicity, electron fraction,
redshift (for an evolving metagalactic UV background), and
temperature.  Tables can be created for any abundance pattern for
elements up to atomic number 30 (Zn) and for any incident radiation
field.  The cooling from H and He is calculated by the non-equilibrium
primordial chemistry and cooling solver.  The contribution of metals
to the cooling is computed within the same subcycling loop as the
coupled primordial chemistry and cooling solver.  The blue curves in
Figure \ref{fig.cooling_rate} show cooling rates calculated with this
method for a gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ and
metallicity of 0.5 Z$_{\odot}$.

\begin{figure}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/cooling_rate.eps}
  \end{center}
  \caption{Radiative cooling rates from the various cooling methods
    available in \enzo.  The black curves show cooling rates from a gas
    with primordial composition using the non-equilibrium chemistry
    network.  The blue curves are for a gas with metallicity of 0.5
    $Z_{\odot}$ computed with the Cloudy cooling method.  The solid
    black and blue lines assume collisional processes only while the
    dashed lines include photo-ionization and photo-heating from a UV
    metagalactic background at $z = 0$ with a gas number density of
    10$^{-4}$ cm$^{-3}$.  The rates shown by the
    dashed lines indicate a net heating below $T \sim 10^{4.5}$, where
    the rapid change in rate is evident (the curve is an absolute
    value so it can be shown on a log plot).  The red curve is the
    tabulated cooling function of \citet{SW87}, which assumes a fully
    ionized gas with metallicity of 0.5 $Z_{\odot}$.}
  \label{fig.cooling_rate}
\end{figure}
